A Mechanised Library for Generic (Co)Programming

### 2024-03-15 paper/paper.tex
edits to section 2

\subsection{Functors and Containers}
\label{sec:containers}
Functors are functions \coq{F : Type -> Type} which additionally have a map
\coq{fmap : forall A B, (A -> B) -> F A -> F B} witnessing the idea that a
functor represents a container for abstract data that can be manipulated
without changing the outer structure. For example, the type 
\coq{List : Type -> Type} is a functor and \coq{List A} is the type of lists
over elements of type \coq{A} with the obvious \coq{fmap} function recursive
over a list and applying the function \coq{A -> B} for each element of the
list. Additionally, the type of lists \coq{List A} arises as the least
fixed-point of the functor \coq{F X = unit + (A * X) }.

@@

\subsection{Types and Morphisms}
\label{sec:types-morphisms}
We address the different forms of equality by defining a class of setoids,
i.e.\ types with na associated equivalence relation, and considering only
functions that respect the associated equivalences, i.e. \emph{proper}
morphisms with respect to the function \emph{respectfulness} relation.  We use
the type-class mechanism, instead of setoids in Coq's standard library, to help
Coq's code extraction mechanism remove any occurence of custom equivalence
relations in the extracted OCaml code.  We use \coq{=e} to denote
the equivalence relation of a setoid. By default, we associate every Coq type
with the standard propositional equality, unless a different equivalence is
specified (i.e.\ we allow overlapping instances, and Coq's propositional
equality takes the lowest priority). 
\begin{coqcode}
Reserved Notation "f =e g" (at level 70, no associativity).
Class setoid A : Type :=
  MkEquiv
    { eqRel : A -> A -> Prop;
      e_refl : forall x, eqRel x x;
      e_sym : forall x y, eqRel x y -> eqRel y x;
      e_trans : forall x y z, eqRel x y -> eqRel y z -> eqRel x z;
    }.
Notation "f =e g" := (eqRel f g).
\end{coqcode}
Given types \coq{A} and  \coq{B}, with their respective
equivalence relations \coq{eA : A -> A -> Prop} and
\coq{eB : B -> B -> Prop}, the we define the type
\coq{A ~> B} to represents \emph{proper} morphisms of the
respectfulness relation of \coq{R} to \coq{R'}.
\begin{coqcode}
Structure morph A {eA : setoid A} B {eB : setoid B} :=
  MkMorph { app :> A -> B; app_eq : forall x y, x =e y -> app x =e app y }.
Notation "A ~> B" := (@morph A _ B _).
\end{coqcode}
Note that we leverage on Coq's type class mechanism to fill in the necessary
equivalence relations. Since Coq's code extraction mechanism will erase any
occurrence of \coq{Prop} in the code, so objects of type \coq{A ~> B} will be
extracted to the OCaml equivalent to \coq{A -> B}. Note the implicit coercion
from \coq{A ~> B} to \coq{A -> B}.
On top of this, we define basic function composition and identity functions:
\begin{coqcode}
Notation "f \o g" = (comp f g).
Definition comp : (B ~> C) ~> (A ~> B) ~> A ~> C := ...
Definition id : A ~> A := ...
\end{coqcode}
Using custom equivalences and proper morphisms, we redefine the definitions of
container extensions and container equality. In particular, container
extensions require a proper morphism to check the validity of positions in
shapes, and container equality now uses equivalences of shapes and contained
elements: 
\begin{coqcode}
Class Cont (Shape : Type) (Pos : Type) := { valid : Shape * Pos ~> bool }.
Inductive AppR `{e : setoid X} (x y : App F X) : Prop :=
  | AppR_ext (Es : shape x =e shape y)
             (Ek : forall e1 e2, val e1 = val e2 -> cont x e1 =e cont y e2).
\end{coqcode}
Note the use of \coq{=e} instead of Coq's standard equality.  For positions,
however, we chose to use Coq's propositional equality, since these would leads
to simpler code. We explain why in Section~\ref{sec:algebras}, and the
mechanisation of initial algebras of container extensions.
%The definitions in
%Figure~\ref{fig:poly} are redefined to account for these changes in a
%straightforward way, i.e. we prove that validity predicates are indeed proper
%morphisms.

This definition leads to the well-known ``setoid hell'', which we mitigate by
providing tactics and notations to automatically discharge proofs of
\coq{app_eq} for morphisms, whenever the types use the standard propositional
equality, or a combination of propositional and extensional equality. However,
our compositional approach allows to build morphisms by plugging in other
morphisms to our combinator, In our framework, our goal is that the
user-provided functions remain small, with relatively straightforward proofs of
\coq{app_eq}.

However, by using this mechanisation, we gain simplified proofs via the use of
Coq's \emph{Generalised Rewriting}.  Since every morphism \coq{f : A ~> B}
satisfies that if \coq{x =e y}, then \coq{f x =e f y}, we can add every
morphism as a proper element of Coq's respectfulness relation. In practice,
this means that we can use the \coq{rewrite} tactic on proofs of type 
\coq{A =e B}, for arbitrary \coq{A} and \coq{B}, whenever they are used as
arguments of morphisms, as well as Coq's \coq{reflexivity}, \coq{symmmetry},
and \coq{transitivity} tactics. For example:
\begin{coqcode}
Goal forall `(f : A ~> B) `(g : B ~> C) `(h : C ~> D) (H : h \o g =e id), 
  h \o (g \o f) =e f.
Proof.  rewrite compA, H.  reflexivity. Qed.
\end{coqcode}

>>>>

\subsection{Functors and Containers}
\label{sec:containers}
Functors are functions \coq{F : Type -> Type} which additionally have a
map \coq{fmap : forall A B, (A -> B) -> F A -> F B} witnessing the idea that a
functor represents a container for abstract data that can be manipulated without
changing the outer structure. For example, the type \coq{List : Type -> Type} is
a functor and \coq{List A} is the type of lists over elements of type \coq{A}
with the obvious \coq{fmap} function recursively traversing a list and applying
the function \coq{A -> B} for each element. Additionally, the type of
lists \coq{List A} arises as the least fixed-point of the
functor \coq{F X = unit + (A * X) }.

@@

\subsection{Types and Morphisms}
\label{sec:types-morphisms}
We address the different forms of equality by defining a class of
\emph{setoids}, types with an associated equivalence relation, and considering
only functions that respect the associated equivalences, or \emph{proper}
morphisms with respect to the function \emph{respectfulness} relation. We use
the type-class mechanism, instead of setoids in Coq's standard library, to help
Coq's code extraction mechanism remove any occurence of custom equivalence
relations in the extracted OCaml code. We use \coq{=e} to denote the equivalence
relation of a setoid. By default, we associate every Coq type with the standard
propositional equality, unless a different equivalence is specified (we allow
overlapping instances, and Coq's propositional equality takes the lowest
priority).
\begin{coqcode}
Reserved Notation "f =e g" (at level 70, no associativity).
Class setoid A : Type :=
  MkEquiv
    { eqRel : A -> A -> Prop;
      e_refl : forall x, eqRel x x;
      e_sym : forall x y, eqRel x y -> eqRel y x;
      e_trans : forall x y z, eqRel x y -> eqRel y z -> eqRel x z;
    }.
Notation "f =e g" := (eqRel f g).
\end{coqcode}
Given types \coq{A} and  \coq{B}, with their respective
equivalence relations \coq{eA : A -> A -> Prop} and
\coq{eB : B -> B -> Prop}, the we define the type
\coq{A ~> B} to represents \emph{proper} morphisms of the
respectfulness relation of \coq{R} to \coq{R'}.
\begin{coqcode}
Structure morph A {eA : setoid A} B {eB : setoid B} :=
  MkMorph { app :> A -> B; app_eq : forall x y, x =e y -> app x =e app y }.
Notation "A ~> B" := (@morph A _ B _).
\end{coqcode}
We rely on Coq's type class mechanism to fill in the necessary equivalence
relations. Coq's code extraction mechanism will erase any occurrence
of \coq{Prop} in the code, so objects of type \coq{A ~> B} will be extracted to
the OCaml equivalent to \coq{A -> B}. Note the implicit coercion
from \coq{A ~> B} to \coq{A -> B}.
On top of this, we define basic function composition and identity functions:
\begin{coqcode}
Notation "f \o g" = (comp f g).
Definition comp : (B ~> C) ~> (A ~> B) ~> A ~> C := ...
Definition id : A ~> A := ...
\end{coqcode}
Using custom equivalences and proper morphisms, we redefine the definitions of
container extensions and container equality. In particular, container
extensions require a proper morphism to check the validity of positions in
shapes, and container equality now uses equivalences of shapes and contained
elements: 
\begin{coqcode}
Class Cont (Shape : Type) (Pos : Type) := { valid : Shape * Pos ~> bool }.
Inductive AppR `{e : setoid X} (x y : App F X) : Prop :=
  | AppR_ext (Es : shape x =e shape y)
             (Ek : forall e1 e2, val e1 = val e2 -> cont x e1 =e cont y e2).
\end{coqcode}
Note the use of \coq{=e} instead of Coq's standard equality.  For positions,
however, we chose to use Coq's propositional equality, since these would leads
to simpler code. We explain why in Section~\ref{sec:algebras}, and the
mechanisation of initial algebras of container extensions.
%The definitions in
%Figure~\ref{fig:poly} are redefined to account for these changes in a
%straightforward way, i.e. we prove that validity predicates are indeed proper
%morphisms.

This definition leads to the well-known ``setoid hell'', which we mitigate by
providing tactics and notations to automatically discharge proofs of
\coq{app_eq} for morphisms, whenever the types use the standard propositional
equality, or a combination of propositional and extensional equality. However,
our compositional approach allows us to build morphisms by plugging in other
morphisms to our combinator. In our framework, our expectation is that the
user-provided functions remain small, with relatively straightforward proofs of
\coq{app_eq}.

However, by using this mechanisation, we gain simplified proofs via the use of
Coq's \emph{Generalised Rewriting}. Since every morphism \coq{f : A ~> B}
satisfies the property that if \coq{x =e y}, then \coq{f x =e f y}, we can add
every morphism as a proper element of Coq's respectfulness relation. In
practice, this means that we can use the \coq{rewrite} tactic on proofs of
type \coq{A =e B}, for arbitrary \coq{A} and \coq{B}, whenever they are used as
arguments of morphisms, as well as Coq's \coq{reflexivity}, \coq{symmmetry},
and \coq{transitivity} tactics. For example:
\begin{coqcode}
Goal forall `(f : A ~> B) `(g : B ~> C) `(h : C ~> D) (H : h \o g =e id), 
  h \o (g \o f) =e f.
Proof.  rewrite compA, H.  reflexivity. Qed.
\end{coqcode}
